Supporting Information:\\Evidence for black-swan events in animal populations

### 2015-07-21 text/ms.tex
Switch to local .bib files
Bankruptcy commit
Major restructuring of text

%!TEX root = anderson-etal-blackswan-timeseries.tex

\section{Abstract}

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, estimate the degree of
heavy-tailedness (presence of black swans) in ecological process noise by
applying a probability model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

\section{Introduction}

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the asteroid collision that may have marked the
end-Cretaceous mass extinction 65 million years ago
\citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with
shifts in mean temperature, are expected to cause the greatest ecological and
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are frequent
events \citep{gaines1993,katz2005}, it remains unclear the extent to which
ecological systems buffer or suffer from black swans \citep{nunez2012}.

There is compelling anecdotal evidence for ecological black swans, but
systematic evidence across taxa has been elusive. A survey of ecologists
indicated that surprising outcomes of field experiments are far more common
than we assume \citep{doak2008}, and events such as the global invasion of
Argentine ants and the mutation of viruses to infect new hosts could be
considered black swans \citep{nunez2012}. In fact, anecdotes of population
catastrophes are numerous and catastrophes may be the most important element
affecting population persistence \citep{mangel1994}. For marine mammal
populations, we have compelling evidence of catastrophes \citep{gerber2001,
  ward2007}, and recently, time-series of marine microbe abundance
\citep{segura2013} and time to extinction for experimental waterflea
populations \citep{drake2014} have been found to follow heavier tailed
distributions than the normal. Despite these examples, as far as we can tell
there have been two systematic surveys for ecological black swans: one with
North American breeding birds and the other with the Global Population
Dynamics Database, which uncovered little clear evidence for them
\citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological
challenges to their detection \citep{allen2001,ward2007}.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not exist in higher taxa. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are normal
tailed on a log scale \citep[e.g.][]{brook2006a,dennis2006,knape2012}.
Alternatively, black-swan dynamics might exist, but our ability to detect
them requires further development of statistical tools. One such tool is the
generalized extreme value distribution, which has been applied to
environmental data \citep[e.g.][]{katz2005}. This distribution describes the
most extreme event per time interval (e.g.~heaviest rainfall per year), but
requires time series that are sufficiently long to be condensed into time
intervals. Another statistical tool involves fitting a catastrophic mixture
distribution in a state space model to quantify the probability that
population events are extreme, although this is also data intensive
\citep{ward2007}. A third tool is to compare the support for fits of thin- and
heavy-tailed distributions \citep{halley2002}, but this analysis did not
quantify the probability of black swans or allow for population dynamics.

Here, we assess the frequency and magnitude of black-swan dynamics across
\NPops\ populations from a wide array of taxonomic groups---mostly birds,
mammals, insects, and fishes. We then identify characteristics of time series
or intrinsic life-history characteristics that are associated with the
detection of black-swan events and attempt to verify known causes. To
accomplish this, we develop a framework for identifying heavy-tailed process
noise in population dynamics, i.e.\ whether the largest stochastic jumps in
abundance from one time step to the next are more extreme than typically seen
with a log-normal distribution. Our framework allows for a range of population
dynamic models, can incorporate observation uncertainty, and can be easily
applied to abundance time series.

\section{Methods}

To obtain estimates of the probability and magnitude of black-swan events, we
fit population dynamic models to abundance time series from around the world.
For each population, we estimated the shape of the process noise tails by
measuring the degrees of freedom ($\nu$) of a Student t-distribution.

\subsection{Time-series data}

We selected abundance time series from the Global Population Dynamics Database
(GPDD; \citeauthor{gpdd2010} \citeyear{gpdd2010}), which contains nearly 5000
time series of abundance from $\sim$1000 species and $\sim$100 taxonomic
orders. We filtered the data (Supporting Information) to remove populations
from less reliable data sources, and those without sufficient data for our
models, and then interpolated some missing values
\citep[\textit{sensu}][]{brook2006a}. Our interpolation affected only
$\sim$\interpPointsPerc \% of the final data points (Table~\ref{tab:stats})
and none of the data points that were later considered black-swan events. Our
final dataset contained \NPops~populations across \NOrders~taxonomic orders
and seven taxonomic classes, with a median of \medianTimeSteps~time steps
(range of \minTimeSteps--\maxTimeSteps) (Table~\ref{tab:stats},
Fig.~\ref{fig:all-ts}).

\subsection{Population models}

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
The growth parameter $\lambda$ represents the expected growth rate if $N_t =
1$. The model is density independent if $b = 1$, maximally density dependent
if $b = 0$, and inversely density dependent if $b < 0$. Usually, the process
noise $\epsilon_t$ is modelled as normally distributed, but in our paper we
assume it is drawn from a t distribution with scale parameter $\sigma$ and
degrees of freedom $\nu$. In previous analyses of the GPDD, the Gompertz was
most often identified as the most parsimonious population model fit to these
data \citep{brook2006}.

By allowing the process noise to be drawn from a Student-t distribution we can
estimate the degree to which the process deviations have heavy tails and are
thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For
example, at $\nu = 3$, the probability of drawing a value more than five
standard deviations below the mean is $0.02$, whereas the probability of
drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$).
As the value of $\nu$ approaches infinity, the distribution approaches the
normal distribution (Fig.~\ref{fig:didactic}a, b). While, across populations,
the extreme process deviations might tend to be more frequently upwards or
downwards events, a small random sample from a heavy-tailed distribution can
have many outliers on one side and appear asymmetric \citep{gelman2013a}.
Therefore, fitting a symmetric t distribution is appropriate for the
relatively short population time series in our dataset and is agnostic towards
the detection of upwards or downwards black swans.

We fit all models in a Bayesian framework using Stan \citep{stan-manual2014}
via the R computing environment \citep{r2014}. Stan samples from the posterior
distribution with an adaptive version of Hamiltonian Markov chain Monte Carlo
called the No-U-Turn Sampler and generally obtains less correlated samples
than algorithms such as the Gibbs sampler \citep{hoffman2014}. We tested to
ensure that the chains had sufficiently converged and that the sampler had
obtained sufficient independent samples from the posterior ($\widehat{R} <
1.05$, $n_\mathrm{eff} > 200$; Supporting Information).

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014};
Figs~\ref{fig:didactic}f--h, \ref{fig:priors}). For $\nu$, we chose an
exponential prior with rate parameter of $0.01$ truncated at values above
two---a slightly less informative prior than suggested by
\citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity. In any case, for $\nu > 20$ the t distribution
is almost indistinguishable from the normal distribution
(Fig.~\ref{fig:didactic}). Based on the shape of the t distribution, we chose
the probability that $\nu < 10$, Pr($\nu < 10$), to define the probability of
heavy-tailed dynamics. When categorizing a population as heavy or normal tailed,
we used a threshold of 0.5 probability.

We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence. Third,
observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance itself (Supporting
Information).

In addition to alternative population models, we investigated the sensitivity
of our results to weaker and stronger priors (exponential rate parameter $=
0.005, 0.02$; Fig.~\ref{fig:priors}, Supporting Information). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Information).

\subsection{Covariates of population dynamic black swans}

We investigated possible covariates of heavy-tailed population dynamics
visually and through multilevel modelling. We plotted characteristics of the
time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
life-history characteristics (body length and maximum lifespan obtained from
\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu <
10)$. We formally investigated these relationships by fitting beta
regression multilevel models \citep{ferrari2004}. The beta probability
distribution can represent continuous response data that range between zero
and one; we fit our models with a logit link as is common for beta and logistic
regression \citep{ferrari2004}. We incorporated standard deviations around the
means for covariates that were derived from Gompertz model parameter
estimates. To account for broad patterns of phylogenetic relatedness, we
allowed for hierarchical intercepts at the taxonomic class, order, and species
level (Supporting Information). We fit our model in Stan with weakly
informative priors on the coefficients \citep{gelman2008d} and variance
parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c}
\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).

Finally, we investigated a sample of populations that our method categorized as
having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where
possible, we found the documented causes of ecological black swans in the
primary data source cited in the GPDD or in other literature describing the
population.

\section{Results}

We found strong, but rare, evidence for black-swan population dynamics. By
defining black-swan dynamics as a greater than $0.5$ probability that $\nu <
10$, our main Gompertz model found evidence for heavy tails most frequently
for birds (\birdPH\%) followed by mammals (\mammalsPH\%), and insects
(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black
swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic
orders. Accounting for time series length and partially pooling inference
across taxonomic class and order with a multilevel model (Supporting
Information), there was stronger evidence for black swans in insect
populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10
orders with the highest median probability of heavy tails were insect
orders---however, there was considerable uncertainty in these estimates
(Fig.~\ref{fig:posteriors}a).

The majority of our heavy-tailed estimates were robust to alternative
population models, observation error, and choice of priors. Our conclusions
were not systematically altered when we included an autocorrelation structure
in the residuals, modelled population growth rates without density dependence,
or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}).
However, setting observation error standard deviation to $0.2$ increased the
median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of
$\baseNuTen$ populations, although the majority of $\nu$ estimates remained
qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$
had little influence on estimates of black-swan dynamics
(Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything,
our models underpredict the true magnitude and probability of heavy tailed
events---especially given the length of the time series in the GPDD
(Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).

Across populations, the probability of observing black-swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Figs~\ref{fig:correlates}, \ref{fig:posteriors}b). Longer time-series length
was the strongest covariate of observing black-swan dynamics. For instance, we
were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a
population with 60 time steps compared to one with 30 time steps. However, the
absolute change in probability with increased time series length was small
($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example,
Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan
events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
majority of documented events (\percBSDown \%) were downward black swans and
involved a combination of multiple factors. For example, a synchronization of
environmental- and predation-mediated population cycles are thought to have
caused a downward black-swan event for a water vole (\textit{Arvicola
  terrestris}) population \citep{saucy1994}. Other black swans were the result
of a sequence of extreme climate events on their own. For instance, severe
winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan
downswings in grey heron (\textit{Ardea cinerea}) abundance in the United
Kingdom \citep{stafford1971}. Our analysis finds that the last event was a
combination of two black-swan events in a row and it took the population three
times longer to recover than predicted \citep{stafford1971}. Downwards black
swans were sometimes followed by upwards black swans. For example, during a
period of population crowding and nest shortages, a population of European
shag cormorants (\textit{Phalacrocorax aristotelis}) on the Farne Islands,
United Kingdom, declined suddenly following a red tide event in 1968
\citep{potts1980}. This freed quality nest sites for first-time breeders,
productivity rapidly increased, and the population experienced a rapid upswing
in abundance. \citep{potts1980}.

\section{Discussion}

We found strong evidence for black swans (heavy-tailed process noise) in
\overallMinPerc--\overallMaxPerc\% of ecological time series. Black swans were
usually (\percBSDown \%) downward events and were detected more frequently in
longer time series and in populations with a smaller magnitude of process
noise. Black swans were not associated with density dependence, population
growth rate, or lifespan. In verified cases, black-swan events were often a
result of the combination of extreme climate, predators and parasite cycles,
and strong changes in human pressures. Our empirical results, sensitivity
analyses, and simulation tests suggest that estimating the tail shape of
process noise is a viable method of detecting black-swan population dynamics
and if anything will underestimate the probability of black-swan events. The
presence of black swans highlights the importance of developing management
strategies that detect quickly, respond to, and are robust to extremes in
population dynamics---particularly as the frequency and magnitude of climatic
extremes increase over the next century \citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. An analysis with an older
version of the GPDD assessed the distribution of abundance time series but
focused on identifying if the log-normal distribution was the most frequently
parsimonious model \citep{halley2002}. For heavy-tailed distributions,
\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
four-parameter Levy stable distribution and found little information criteria
support for these distributions in longer time series. However, black-swan
events are by definition rare, and the majority of time series in the GPDD are
short for these purposes. Therefore, we would not expect to observe black
swans in a large proportion of populations. By quantifying the probability of
heavy-tails and allowing for non-stationary time series and density
dependence, our analysis allows for a more nuanced description of the evidence
of ecological black swans. In an earlier study, \citet{keitt1998} described
heavy (power law) tails in breeding bird population abundance. However, this
finding was challenged by \citet{allen2001}, who showed that mixing data
across species could falsely generate heavy tails.
%This rebuttal supports our
%use of the t distribution to represent heavy tails, since a t distribution can
%be represented by a mixture of normal distributions \citep[with the same mean
%and inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black-swan dynamics in ecological time series
because of unmodelled intrinsic properties of populations or extrinsic forces
acting on populations. Since a t-distribution can be formed by a mixture of
normal distributions \citep[with the same mean and inverse-gamma-distributed
variances,][]{gelman2014}, we could observe heavy tails if we miss some
underlying mixture of intrinsic processes \citep{allen2001}. That process
might be an aggregation of populations across space, or population diversity,
or an intrinsic change in population variability through time. Extrinsic
forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These
forces could be extreme themselves. For example, extreme climate, predation
from (or competition with) other species experiencing black swans, or sharp
changes in human pressure such hunting, fishing, or habitat destruction might
cause black swans. Alternatively, the synchrony of multiple ``normal''
extrinsic forces could give rise to black-swan ecological dynamics. This could
occur if with a synergistic interaction \citep[e.g.][]{kirby2009} or even if
non-synergistic forces experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results.
The GPDD data represent a taxonomically and geographically biased sample of
populations---the longer time series we focus on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to
find qualitatively similar evidence for black swans in many other large
taxonomic or geographic samples of populations, the common forces driving
those black swans would likely differ (e.g.~severe winters in
Table~\ref{tab:sparks}). In addition to a possibly biased sample of
populations, some black-swan detections could just be recording mistakes, or
conversely, some extreme observations may have been discarded or altered if
they were erroneously suspected of being recording mistakes. Indeed, we
discarded three of the populations that our method initially identified as
heavy tailed because they turned out to be data-entry errors (Supporting
Information). A final caveat is that the temporal scales of observation and
population dynamics vary considerably across populations in the GPDD and this
likely influences the detection of heavy tails. As an example, if we make
frequent observations relative to generation time (e.g.~for many large-bodied
mammals) we will average across generations and perhaps miss black swans.
Conversely, if we census populations infrequently relative to generation time
(e.g.~many insects in the GPDD) the recorded data may average across extreme
and less-extreme events and also dampen black-swan dynamics.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of
policy directions. Ecological resource management can draw from other
disciplines that focus on heavy tails. For example, earthquake preparedness
and response is focussed on black-swan events. Similarly to ecological black
swans, we can rarely predict the specific location and timing of large
earthquakes. But, earthquake preparedness involves spatial planning based on
forecast probabilities to focus early detection efforts and develop disaster
response plans. A similar focus might benefit resource management once our
ability to predict the spatial probability and covariates of ecological black
swans improves. The presence of black swans also suggests that we develop
management policy that is robust to heavy tails. For instance, setting target
population abundances that are appropriately set back from critical limits may
buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic,
phenotypic, and behavioural diversity may allow some components of populations
to persist when others are affected by disease or extreme environmental forces
\citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and
unexpected, surprising, or counterintuitive ecological dynamics offer a
tremendous opportunity to learn about ecological systems, evaluate when our
models break down, and adjust future management policy \citep{doak2008,
  pine-iii2009, lindenmayer2010}.

Our results suggest a number of research questions related to ecological black
swans. Given that black swans do occur, can we forecast the probability of
black swans in space and time? Furthermore, what management policies allow us
to detect them quickly after they happen? Can we isolate the components of
ecological dynamics that experience black swans by moving from
phenomenological models such as the Gompertz to mechanistic models that, for
example, take into account recruitment dynamics? We expect that greater
insight into the mechanisms and covariates of ecological black swans may be
best obtained through specific geographic and taxonomic subsets of data where
longer time series with low levels of observation error are available
\citep[e.g.][]{segura2013}.

Most importantly, what is the impact of allowing for black swans in forecasts
of ecological risk? Even extremely rare catastrophes can have a profound
influence on population persistence \citep{mangel1994}. In recent decades,
ecology has moved toward focussing on aspects of variance in addition to mean
responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that
an added focus on ecological extremes represents the next frontier,
particularly in the face of increased climate extremes \citep{meehl2004,
  thompson2013, ipcc2012}. Financial analysts are concerned with the shape of
the downward tails of financial returns because these directly impact
estimates of risk---the probability of a specific magnitude of undesired event
occurring \citep{rachev2008}. A comparable focus in ecology would increase our
estimates of extinction risk, since these would be disproportionately impacted
by downward black-swan events.

\section{Acknowledgements}

We thank J.D. Yeakel and other members of the Earth to Ocean research
group for helpful discussions and comments. We are grateful to the
contributors and maintainers of the Global Population Dynamics Database and to
Compute Canada's WestGrid high-performance computing resources. Silhouette
images were obtained from \texttt{phylopic.org} under Creative Commons
licenses; sources are listed in the Supporting Information. Funding was
provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural
Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada
Research Chairs Program (NKD).

\section{Supporting Information}

The following supporting information is available online for this article:\\
Tables S1 and S2\\
Figures S1--S6\\
Example Stan code for a heavy-tailed Gompertz model and multilevel beta
regression\\
The GPDD IDs used in our analysis\\
All code and data to recreate our analysis are available at:\\
\url{https://github.com/seananderson/heavy-tails}

\renewcommand{\baselinestretch}{\tighttextstretch}
\normalsize
\bibliographystyle{apalike}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

\clearpage
\renewcommand{\baselinestretch}{\textstretch}
\normalsize

\section{Tables}

\input{sparks.tex}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg2.pdf}
\caption[An illustration of fitting population dynamic models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$.]{
An illustration of fitting population dynamic models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value more than five
standard deviations below the mean is 1.8\%, whereas the probability of drawing
such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%).
(c--e) Simulated population dynamics from a Gompertz model with process noise
drawn from t distributions with three different values of $\nu$. Coloured dots
in panels c and d represent jumps with less than a 1 in 1000 chance of
occurring in a normal distribution. (f--h) Estimates of $\nu$ from models fit
to the times series in panels c--e. Shown are the posterior samples
(histograms), median and interquartile range of the posterior (IQR) (dots and
line segments), and the exponential prior on $\nu$ (dashed lines). Colour
shading behind panels f--h illustrates the region of heavy tails.
}
\label{fig:didactic}
\end{center}
\end{figure}

>>>>

\textbf{Title: Evidence for black-swan events in animal populations}

\textbf{Authors}: Sean C. Anderson\textsuperscript{1,2*}, Trevor A. Branch\textsuperscript{3}, Andrew B. Cooper\textsuperscript{2}, Nicholas K. Dulvy\textsuperscript{1}

\textsuperscript{1}Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada

\textsuperscript{2}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada

\textsuperscript{3}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA

\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sean\_anderson@sfu.ca

\textbf{Abstract}:

Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. We derived a model for estimating the frequency of black-swan events and applied it to abundance time series of 609 animal populations. We find strong evidence for black-swan events, but they are rare, occurring in 3--5\% of populations: most frequently for birds (4--8\%), mammals (4--6\%), and insects (2--3\%). Black swan events were predominantly (87\%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases. Black swan events are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate or severe winters. Data and simulations reveal that black swan-events are more likely to be detected in longer and less noisy time series. Our results show that accurate forecasts of extinction risk require accounting for the possibility of black swans and the predominance of crashes over increases, or predicted abundance will underestimate the probability of crashes by a factor of XX--XX. In addition, the presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.

@@

Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, may2008}. Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the most damaging earthquakes greatly exceed their expected frequency\cite{sornette2009}. In ecology, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}. Indeed, catastrophes may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}. Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, dennis2006, knape2012}. If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.

Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes. We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes. To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in abundance from one time step to the next are more extreme than typically seen with a lognormal distribution. Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution\emph{.} By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1). For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is \(0.02\), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^-7\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2, Table~S2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, we found stronger evidence for black swans in insect populations than is apparent in Fig.~2---four of eight orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~3a).

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Fig.~S3). Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Fig.~S3). The strength of the prior on ν had little influence on estimates of black-swan dynamics (Fig.~S4). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Figs S5, S6).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Figs 3b, 4). Longer time-series length was the strongest covariate of observing black-swan dynamics. For instance, the expected probability density suggesting black-swan dynamics was about 1.6 times greater for a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Fig.~4).

We examined all time series with published explanations of why black-swan events occurred (Table~S2). The majority (87\%) were downward black swans and involved a combination of multiple factors. For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted\cite{stafford1971}. Downward black swans were sometimes followed by upward black swans. For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom, declined suddenly following a red tide event in 1968\cite{potts1980}. This freed quality nest sites for first-time breeders, productivity rapidly increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.

Given the prevalence of downwards events, we refit our heavy-tailed models estimating the skewness of process noise (\(\gamma\); Figs 1a, SX) and used these models to make near-term risk forecasts. Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, SX). In contrast, populations with little evidence of heavy tails (median \(\nu \geq 70\)) also had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these heavy-tailed populations forward five years revealed that assuming lognormal process noise underestimated risk by X--X fold (ratio of interquartile range of 99\% lower credible interval of abundance). \emph{TODO: closing sentence emphasizing the implications of this?}

Thus our analysis provides strong evidence for the presence of downwardly skewed black-swan events in ecological process noise of higher taxa. Unmodelled intrinsic properties of populations or extrinsic forces acting on populations could generate black-swan dynamics. Given that a mixture of normal distributions with different variances can generate a Student-t distribution with heavy tails\cite{allen2001}, we could observe black-swan dynamics if we miss an underlying mixture of processes. Those processes might be an aggregation of populations across space, population diversity\cite{anderson2015}, or an intrinsic change in population variability through time. Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012}. These forces could be extreme themselves. For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction might cause black-swan events. Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan ecological dynamics. This could occur with synergistic interactions\cite{kirby2009} or even if non-synergistic forces experience a rare alignment\cite{denny2009}.

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ. Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supporting Information). Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}. A similar focus might benefit ecological resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{hilborn2003, schindler2010, anderson2015}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, pine-iii2009, lindenmayer2010}

Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}. In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}. Financial analysts are concerned with the shape of the downward tails of financial returns because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events. Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.
